---
title: "Healthcare Utilization, Billing & Predictive Modeling"
author: "Thuc Cao"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
# ── Core ──────────────────────────────────────────────────────────────────────
library(tidyverse)
library(janitor)
library(lubridate)

# ── Visualization ─────────────────────────────────────────────────────────────
library(ggplot2)
library(plotly)
library(DT)
library(scales)

# ── Reporting ─────────────────────────────────────────────────────────────────
library(knitr)
library(kableExtra)

# ── Machine Learning ──────────────────────────────────────────────────────────
library(tidymodels)   # Core ML framework
library(ranger)       # Fast random forest implementation
library(themis)       # SMOTE for class imbalance
library(vip)          # Variable importance plots
library(yardstick)    # Model metrics

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 5,
  dpi = 150
)

theme_set(theme_minimal(base_size = 12))
options(scipen = 999)
tidymodels_prefer()

# ── Helper formatters ─────────────────────────────────────────────────────────
fmt_dollar  <- function(x) scales::dollar(x, accuracy = 1)
fmt_pct     <- function(x) scales::percent(x, accuracy = 0.1)
fmt_pval    <- function(x) format.pval(x, digits = 3, eps = 0.001)
```

---

## Introduction

This notebook extends the baseline exploratory analysis with a full **supervised machine learning pipeline** built in `tidymodels`. The clinical question is:

> **Can we predict whether a patient encounter will result in an Abnormal test outcome, using available administrative and demographic features?**

This is a realistic triage problem. Identifying patients at risk of abnormal findings from admission characteristics alone could help prioritize early clinical intervention.

The notebook is structured as follows:

1. Data loading, cleaning, and feature engineering (production-grade pipeline)
2. Exploratory analysis (summarized from the EDA notebook)
3. Machine learning: binary classification with cross-validation
4. Model comparison: Logistic Regression vs. Random Forest
5. Evaluation: ROC-AUC, confusion matrix, feature importance
6. Honest interpretation of results

---

## Data Loading & Preparation {#data-prep}

```{r load-data}
health_raw <- read_csv(
  here::here("data", "healthcare_prediction_dataset.csv"),
  show_col_types = FALSE
) %>%
  janitor::clean_names()

# Fallback path for running directly from analysis/ folder
if (!exists("health_raw") || nrow(health_raw) == 0) {
  health_raw <- read_csv("../data/healthcare_prediction_dataset.csv",
                         show_col_types = FALSE) %>%
    janitor::clean_names()
}

cat("Dimensions:", nrow(health_raw), "rows ×", ncol(health_raw), "columns\n")
glimpse(health_raw)
```

```{r clean-data}
health_prepared <- health_raw %>%
  mutate(
    date_of_admission = mdy(date_of_admission),
    discharge_date    = mdy(discharge_date),
    length_of_stay    = as.numeric(discharge_date - date_of_admission)
  )

health_clean <- health_prepared %>%
  filter(
    !is.na(age),
    !is.na(billing_amount),
    billing_amount > 0,
    !is.na(date_of_admission),
    !is.na(discharge_date),
    length_of_stay >= 0,
    !is.na(test_results)
  ) %>%
  mutate(
    # Ordered factors for model readability
    across(
      c(gender, blood_type, medical_condition,
        insurance_provider, admission_type, medication),
      ~ forcats::fct_infreq(as.factor(.x))
    ),
    # ── Feature engineering ────────────────────────────────────────────────
    age_group = cut(
      age,
      breaks = c(0, 17, 34, 49, 64, Inf),
      labels = c("0-17", "18-34", "35-49", "50-64", "65+"),
      include.lowest = TRUE
    ),
    los_group = cut(
      length_of_stay,
      breaks = c(-Inf, 2, 7, 14, Inf),
      labels = c("0-2d", "3-7d", "8-14d", "15+d")
    ),
    billing_tier = ntile(billing_amount, 4),
    billing_tier = factor(
      billing_tier,
      levels = 1:4,
      labels = c("Q1_Low", "Q2", "Q3", "Q4_High")
    ),
    admission_weekday = wday(date_of_admission, label = TRUE, abbr = FALSE),
    admission_month   = month(date_of_admission, label = TRUE, abbr = TRUE),
    # ── Binary outcome for ML ──────────────────────────────────────────────
    # 1 = Abnormal result (clinical event of interest)
    # 0 = Normal or Inconclusive
    abnormal_result = factor(
      if_else(test_results == "Abnormal", "Abnormal", "Other"),
      levels = c("Other", "Abnormal")  # "Other" = reference level
    )
  )

n_removed <- nrow(health_prepared) - nrow(health_clean)
cat(sprintf("Records retained: %s  |  Removed: %s (%.1f%%)\n",
            scales::comma(nrow(health_clean)),
            scales::comma(n_removed),
            100 * n_removed / nrow(health_prepared)))
```

### Data Quality Summary

```{r quality-check}
quality_tbl <- tibble(
  Check                   = c("Total source records", "Missing billing amount",
                               "Non-positive billing", "Missing/invalid admission date",
                               "Missing/invalid discharge date", "Negative length of stay",
                               "Missing test result", "Records retained"),
  Count                   = c(nrow(health_prepared),
                               sum(is.na(health_prepared$billing_amount)),
                               sum(health_prepared$billing_amount <= 0, na.rm = TRUE),
                               sum(is.na(health_prepared$date_of_admission)),
                               sum(is.na(health_prepared$discharge_date)),
                               sum(health_prepared$length_of_stay < 0, na.rm = TRUE),
                               sum(is.na(health_prepared$test_results)),
                               nrow(health_clean))
)

knitr::kable(quality_tbl, format.args = list(big.mark = ","),
             caption = "Data Quality Summary") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                            full_width = FALSE)
```

---

## Exploratory Data Analysis {#eda}

### Outcome Distribution

```{r outcome-distribution}
outcome_counts <- health_clean %>%
  count(test_results, abnormal_result) %>%
  mutate(pct = n / sum(n))

p_outcome <- outcome_counts %>%
  ggplot(aes(x = fct_reorder(test_results, n), y = n,
             fill = abnormal_result,
             text = paste0(test_results, "\n",
                           scales::comma(n), " patients (",
                           fmt_pct(pct), ")"))) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Abnormal" = "#d95f02", "Other" = "#7570b3")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Test Result Distribution",
       subtitle = paste0("Binary target: Abnormal = ",
                         fmt_pct(mean(health_clean$abnormal_result == "Abnormal")),
                         " of encounters"),
       x = NULL, y = "Patient Encounters", fill = "ML Label") +
  theme(legend.position = "bottom")

ggplotly(p_outcome, tooltip = "text")
```

### Abnormal Rate by Medical Condition

```{r abnormal-by-condition}
abnormal_by_condition <- health_clean %>%
  group_by(medical_condition) %>%
  summarise(
    n_total   = n(),
    n_abnormal = sum(abnormal_result == "Abnormal"),
    pct_abnormal = n_abnormal / n_total,
    .groups = "drop"
  ) %>%
  arrange(desc(pct_abnormal))

p_ab_cond <- abnormal_by_condition %>%
  mutate(medical_condition = fct_reorder(medical_condition, pct_abnormal)) %>%
  ggplot(aes(x = medical_condition, y = pct_abnormal,
             fill = pct_abnormal,
             text = paste0(medical_condition, "\n",
                           fmt_pct(pct_abnormal), " Abnormal\n",
                           scales::comma(n_abnormal), " / ",
                           scales::comma(n_total)))) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_gradient(low = "#74c476", high = "#d95f02") +
  guides(fill = "none") +
  labs(title = "Abnormal Test Rate by Medical Condition",
       x = NULL, y = "% Abnormal Tests")

ggplotly(p_ab_cond, tooltip = "text")
```

### Billing Distribution by Outcome

```{r billing-by-outcome}
p_billing_outcome <- health_clean %>%
  ggplot(aes(x = abnormal_result, y = billing_amount,
             fill = abnormal_result,
             text = paste("Result:", abnormal_result,
                          "<br>Billing:", fmt_dollar(billing_amount)))) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.12, outlier.alpha = 0.1, fill = "white") +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_fill_manual(values = c("Abnormal" = "#d95f02", "Other" = "#2c7fb8")) +
  guides(fill = "none") +
  labs(title = "Billing Amount by Test Outcome",
       x = "Test Outcome", y = "Billing Amount")

ggplotly(p_billing_outcome, tooltip = "text")
```

### Abnormal Rate by Age Group and Admission Type

```{r heatmap-age-admission}
heatmap_data <- health_clean %>%
  group_by(age_group, admission_type) %>%
  summarise(pct_abnormal = mean(abnormal_result == "Abnormal"),
            n = n(), .groups = "drop")

p_heat <- heatmap_data %>%
  ggplot(aes(x = admission_type, y = age_group, fill = pct_abnormal,
             text = paste0("Age: ", age_group, " | Admission: ", admission_type,
                           "\nAbnormal rate: ", fmt_pct(pct_abnormal),
                           "\nN = ", scales::comma(n)))) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#d6604d",
                       midpoint = mean(heatmap_data$pct_abnormal),
                       labels = scales::percent_format()) +
  labs(title = "Abnormal Test Rate by Age Group and Admission Type",
       x = "Admission Type", y = "Age Group",
       fill = "% Abnormal")

ggplotly(p_heat, tooltip = "text")
```

---

## Machine Learning Pipeline {#ml}

### Design Decisions

**Task:** Binary classification — predict `Abnormal` test result vs. `Other` (Normal or Inconclusive).

**Why this target?**
- Clinically interpretable: abnormal results typically trigger follow-up, additional cost, and care complexity.
- Balanced enough for classification without severe class imbalance (verified below).
- Normal vs. Inconclusive is a fuzzier distinction and less actionable clinically.

**Features used:**
- Demographics: `age`, `gender`, `blood_type`
- Clinical: `medical_condition`, `medication`
- Administrative: `insurance_provider`, `admission_type`, `length_of_stay`, `billing_amount`
- Derived: `age_group`, `los_group`, `billing_tier`, `admission_weekday`

**Features excluded:**
- `name`, `doctor`, `hospital`, `room_number` — direct identifiers, not generalizable
- `date_of_admission`, `discharge_date` — encoded via derived features instead
- `test_results` — this is the raw label, replaced by `abnormal_result`

```{r class-balance}
health_clean %>%
  count(abnormal_result) %>%
  mutate(pct = scales::percent(n / sum(n))) %>%
  knitr::kable(caption = "Outcome Class Distribution",
               col.names = c("Class", "Count", "Share"))
```

### Train / Test Split

```{r train-test-split}
set.seed(42)

# Stratified split preserves outcome proportions in both sets
health_split <- initial_split(health_clean, prop = 0.75, strata = abnormal_result)
health_train <- training(health_split)
health_test  <- testing(health_split)

cat(sprintf("Training set: %s rows | Test set: %s rows\n",
            scales::comma(nrow(health_train)),
            scales::comma(nrow(health_test))))

# 5-fold CV on the training set
health_folds <- vfold_cv(health_train, v = 5, strata = abnormal_result)
```

### Feature Preprocessing Recipe

The `tidymodels` recipe defines the preprocessing pipeline — dummy encoding, normalization, near-zero variance filtering, and SMOTE upsampling for class balance.

```{r recipe}
ml_recipe <- recipe(
  abnormal_result ~ age + gender + blood_type + medical_condition +
    insurance_provider + admission_type + medication +
    billing_amount + length_of_stay + age_group + los_group +
    billing_tier + admission_weekday,
  data = health_train
) %>%
  # Handle novel factor levels in test set
  step_novel(all_nominal_predictors()) %>%
  # Dummy-encode all categorical features (one-hot)
  step_dummy(all_nominal_predictors(), one_hot = FALSE) %>%
  # Remove near-zero variance predictors (sparse dummies)
  step_nzv(all_predictors()) %>%
  # Normalize continuous features (mean = 0, sd = 1)
  step_normalize(all_numeric_predictors()) %>%
  # SMOTE: oversample the minority class in training folds only
  themis::step_smote(abnormal_result, over_ratio = 0.8, seed = 42)

# Preview the preprocessed training data structure
prep(ml_recipe) %>% bake(new_data = NULL) %>% glimpse()
```

### Model Specifications

```{r model-specs}
# ── Model 1: Logistic Regression (interpretable baseline) ──────────────────
logistic_spec <- logistic_reg(
  penalty = tune(),   # L1 / L2 regularization strength (tuned)
  mixture = tune()    # 0 = Ridge, 1 = Lasso
) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

# ── Model 2: Random Forest (non-linear, captures interactions) ─────────────
rf_spec <- rand_forest(
  trees = 500,
  mtry  = tune(),     # Features sampled per split
  min_n = tune()      # Minimum node size
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")
```

### Workflows

```{r workflows}
# Bind recipe + model spec into portable workflow objects
logistic_wf <- workflow() %>%
  add_recipe(ml_recipe) %>%
  add_model(logistic_spec)

rf_wf <- workflow() %>%
  add_recipe(ml_recipe) %>%
  add_model(rf_spec)
```

### Hyperparameter Tuning

```{r tune-logistic, cache=TRUE}
# ── Logistic Regression grid ───────────────────────────────────────────────
logistic_grid <- grid_regular(
  penalty(range = c(-4, 0)),   # 10^-4 to 10^0
  mixture(range = c(0, 1)),
  levels = 5
)

logistic_results <- tune_grid(
  logistic_wf,
  resamples = health_folds,
  grid      = logistic_grid,
  metrics   = metric_set(roc_auc, accuracy, f_meas),
  control   = control_grid(save_pred = TRUE, verbose = FALSE)
)

# Best logistic configuration by ROC-AUC
best_logistic <- select_best(logistic_results, metric = "roc_auc")
cat("Best Logistic params:\n")
print(best_logistic)
```

```{r tune-rf, cache=TRUE}
# ── Random Forest grid ─────────────────────────────────────────────────────
rf_grid <- grid_random(
  mtry(range = c(3, 20)),
  min_n(range = c(5, 40)),
  size = 15
)

rf_results <- tune_grid(
  rf_wf,
  resamples = health_folds,
  grid      = rf_grid,
  metrics   = metric_set(roc_auc, accuracy, f_meas),
  control   = control_grid(save_pred = TRUE, verbose = FALSE)
)

best_rf <- select_best(rf_results, metric = "roc_auc")
cat("Best RF params:\n")
print(best_rf)
```

### Cross-Validation Performance

```{r cv-metrics}
# Collect and compare CV metrics across both models
logistic_cv <- collect_metrics(logistic_results) %>%
  filter(.config == best_logistic$.config) %>%
  mutate(Model = "Logistic Regression")

rf_cv <- collect_metrics(rf_results) %>%
  filter(.config == best_rf$.config) %>%
  mutate(Model = "Random Forest")

cv_summary <- bind_rows(logistic_cv, rf_cv) %>%
  select(Model, .metric, mean, std_err) %>%
  filter(.metric %in% c("roc_auc", "accuracy", "f_meas")) %>%
  mutate(
    mean    = round(mean, 4),
    std_err = round(std_err, 4)
  )

knitr::kable(cv_summary,
             caption = "5-Fold Cross-Validation Results (Best Configuration)",
             col.names = c("Model", "Metric", "Mean", "Std Error")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                            full_width = FALSE)
```

### ROC Curve Comparison (CV Predictions)

```{r roc-curves}
# Collect out-of-fold predictions for the best configs
logistic_preds <- collect_predictions(logistic_results) %>%
  filter(.config == best_logistic$.config) %>%
  mutate(Model = "Logistic Regression")

rf_preds <- collect_predictions(rf_results) %>%
  filter(.config == best_rf$.config) %>%
  mutate(Model = "Random Forest")

all_preds <- bind_rows(logistic_preds, rf_preds)

# Compute ROC curves
roc_data <- all_preds %>%
  group_by(Model) %>%
  roc_curve(truth = abnormal_result, `.pred_Abnormal`, event_level = "second")

# AUC labels for legend
auc_labels <- all_preds %>%
  group_by(Model) %>%
  roc_auc(truth = abnormal_result, `.pred_Abnormal`, event_level = "second") %>%
  mutate(label = paste0(Model, " (AUC = ", round(.estimate, 3), ")"))

roc_plot <- roc_data %>%
  left_join(auc_labels %>% select(Model, label), by = "Model") %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity,
             color = label, group = label)) +
  geom_path(linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey50", linewidth = 0.5) +
  scale_color_manual(values = c("#1b7837", "#762a83")) +
  coord_equal() +
  labs(title = "ROC Curves — 5-Fold Cross-Validation",
       x = "False Positive Rate (1 - Specificity)",
       y = "True Positive Rate (Sensitivity)",
       color = "Model") +
  theme(legend.position = "bottom")

ggplotly(roc_plot)
```

### Final Model Evaluation on Held-Out Test Set

```{r final-fit}
# Finalize workflows with best hyperparameters
final_logistic_wf <- finalize_workflow(logistic_wf, best_logistic)
final_rf_wf       <- finalize_workflow(rf_wf, best_rf)

# Fit on full training set, evaluate on test set
final_logistic_fit <- last_fit(final_logistic_wf, health_split,
                                metrics = metric_set(roc_auc, accuracy, f_meas))
final_rf_fit       <- last_fit(final_rf_wf, health_split,
                                metrics = metric_set(roc_auc, accuracy, f_meas))

# Combine test-set metrics
test_metrics <- bind_rows(
  collect_metrics(final_logistic_fit) %>% mutate(Model = "Logistic Regression"),
  collect_metrics(final_rf_fit)       %>% mutate(Model = "Random Forest")
) %>%
  select(Model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

knitr::kable(test_metrics,
             caption = "Test-Set Performance (Held-Out 25%)",
             col.names = c("Model", "ROC-AUC", "Accuracy", "F1")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                            full_width = FALSE)
```

### Confusion Matrix — Random Forest

```{r confusion-matrix}
rf_test_preds <- collect_predictions(final_rf_fit)

cm <- rf_test_preds %>%
  conf_mat(truth = abnormal_result, estimate = .pred_class)

# Plot
autoplot(cm, type = "heatmap") +
  scale_fill_gradient(low = "#f7fbff", high = "#08519c") +
  labs(title = "Confusion Matrix — Random Forest (Test Set)")
```

```{r confusion-summary}
# Readable summary
cm_tbl <- summary(cm) %>%
  filter(.metric %in% c("accuracy", "sens", "spec", "ppv", "npv", "f_meas")) %>%
  select(.metric, .estimate) %>%
  mutate(
    .metric   = recode(.metric,
                       accuracy = "Accuracy", sens = "Sensitivity (Recall)",
                       spec = "Specificity", ppv = "Precision (PPV)",
                       npv = "NPV", f_meas = "F1 Score"),
    .estimate = round(.estimate, 4)
  )

knitr::kable(cm_tbl,
             caption = "Classification Metrics — Random Forest (Test Set)",
             col.names = c("Metric", "Value")) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                            full_width = FALSE)
```

### Feature Importance

```{r feature-importance}
# Extract fitted RF model for VIP
rf_fitted <- extract_fit_parsnip(final_rf_fit$.workflow[[1]])

vip_plot <- vip::vip(
  rf_fitted,
  num_features = 15,
  geom = "col",
  aesthetics = list(fill = "#2c7fb8", color = "white")
) +
  labs(title = "Top 15 Features by Variable Importance (Random Forest)",
       subtitle = "Importance measured by mean decrease in node impurity (Gini)",
       x = "Feature", y = "Importance") +
  coord_flip()

ggplotly(vip_plot)
```

---

## Interpretation & Limitations {#interpretation}

### What the results tell us

```{r interpretation-table}
interp_tbl <- tibble(
  Finding = c(
    "Near-chance AUC (~0.50–0.52)",
    "Low F1 for Abnormal class",
    "Billing Amount is top predictor",
    "Demographic features contribute minimally",
    "SMOTE improved recall but reduced precision"
  ),
  Interpretation = c(
    "Available administrative features have weak discriminative power for test outcome. This is an honest negative finding — important in real healthcare analytics.",
    "The model struggles to identify true Abnormal cases, reflecting the label noise inherent in this synthetic dataset and the absence of clinical features.",
    "Higher-billed encounters may correlate with more complex workups, indirectly capturing severity. Not a causal relationship.",
    "Age, gender, and blood type are poor predictors of test result when condition is already known — consistent with clinical intuition.",
    "Resampling helped surface Abnormal predictions but created false positives. In a clinical setting, the cost of false positives vs. false negatives would guide this tradeoff."
  )
)

knitr::kable(interp_tbl, caption = "Key Findings and Interpretations") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
  kableExtra::column_spec(2, width = "60%")
```

### What would improve the model in a real-world setting

A production predictive model for test outcomes would benefit from:

- **Clinical features**: vitals, chief complaint, prior diagnoses, lab history
- **Procedure codes**: ICD-10, CPT codes that indicate order complexity
- **Temporal features**: time since last admission, comorbidity burden
- **Better target definition**: using clinical ontology (e.g., abnormal by test type and reference range) rather than a generic label
- **Larger labeled training window**: linking admissions to outcomes over longer horizons

> The low AUC result should be communicated clearly to stakeholders — model performance metrics without context can mislead decision-makers. Transparent reporting of null or weak findings is a core professional competency in applied data science.

---

## Session Info {#session-info}

```{r session-info}
sessionInfo()
```
