---
title: "Healthcare Utilization and Billing Analysis"
author: "Thuc Cao"
output:
  html_notebook:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
# Load packages for data preparation, visualization, interactive tables, and reporting.
library(tidyverse)
library(knitr)
library(DT)
library(plotly)
library(janitor)
library(lubridate)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5
)

theme_set(theme_minimal(base_size = 12))
options(scipen = 999)

format_currency <- function(x) scales::dollar(x, accuracy = 1)
format_percent <- function(x) scales::percent(x, accuracy = 0.1)
format_p_value <- function(x) format.pval(x, digits = 3, eps = 0.001)
```

## Introduction

Healthcare data analysis helps decision makers understand patient populations, service utilization, and potential cost drivers. This report analyzes the `healthcare_prediction_dataset.csv` file to explore billing patterns, admission types, patient demographics, medical conditions, insurance providers, and length of stay.

The goal is to produce a professional exploratory analysis that answers four practical questions:

- What does the cleaned patient encounter dataset look like?
- Which conditions, insurers, and admission types account for the largest patient volumes and billing differences?
- How do age and length of stay relate to billing amount?
- Are medical condition and admission type statistically associated?

This analysis is exploratory. It identifies patterns and associations, but it should not be interpreted as a causal or clinical decision model.

## 1. Data Loading and Preparation

The first step is to load the dataset, standardize column names, convert dates, and create a few derived fields that make the analysis easier to interpret.

```{r load-data}
health_raw <- read_csv("healthcare_prediction_dataset.csv", show_col_types = FALSE) %>%
  janitor::clean_names()

glimpse(health_raw)
```

The dataset includes patient-level fields such as demographics, condition, admission date, discharge date, insurance provider, billing amount, medication, and test result.

```{r prepare-data}
health_prepared <- health_raw %>%
  mutate(
    date_of_admission = mdy(date_of_admission),
    discharge_date = mdy(discharge_date),
    length_of_stay = as.numeric(discharge_date - date_of_admission)
  )

health_clean <- health_prepared %>%
  filter(
    !is.na(age),
    !is.na(billing_amount),
    billing_amount > 0,
    !is.na(date_of_admission),
    !is.na(discharge_date),
    length_of_stay >= 0
  ) %>%
  mutate(
    across(
      c(gender, blood_type, medical_condition, insurance_provider,
        admission_type, medication, test_results),
      ~ forcats::fct_infreq(as.factor(.x))
    )
  )

removed_records <- nrow(health_prepared) - nrow(health_clean)

glimpse(health_clean)
```

The cleaning step removes records with missing critical fields, invalid dates, negative length of stay, or non-positive billing amounts. In this dataset, `r scales::comma(removed_records)` records were excluded from the analytical sample.

## 2. Data Quality Review

A professional healthcare analysis should make the cleaning logic transparent. The table below summarizes the main quality checks before final filtering.

```{r data-quality-checks}
quality_checks <- tibble(
  Check = c(
    "Total source records",
    "Missing billing amount",
    "Non-positive billing amount",
    "Missing or invalid admission date",
    "Missing or invalid discharge date",
    "Negative length of stay",
    "Records retained for analysis"
  ),
  Count = c(
    nrow(health_prepared),
    sum(is.na(health_prepared$billing_amount)),
    sum(health_prepared$billing_amount <= 0, na.rm = TRUE),
    sum(is.na(health_prepared$date_of_admission)),
    sum(is.na(health_prepared$discharge_date)),
    sum(health_prepared$length_of_stay < 0, na.rm = TRUE),
    nrow(health_clean)
  )
)

knitr::kable(quality_checks, caption = "Data Quality Checks")
```

```{r missing-values}
missing_tbl <- health_raw %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "Field", values_to = "Missing_Count") %>%
  mutate(Missing_Rate = Missing_Count / nrow(health_raw)) %>%
  arrange(desc(Missing_Count))

knitr::kable(
  missing_tbl %>%
    mutate(Missing_Rate = format_percent(Missing_Rate)),
  caption = "Missing Values by Field"
)
```

For privacy-conscious reporting, the preview below excludes direct identifiers such as patient name, doctor, hospital, and room number. Even when a dataset is synthetic, a portfolio healthcare analysis should model careful handling of patient-level fields.

```{r deidentified-preview}
preview_tbl <- health_clean %>%
  transmute(
    Age = age,
    Gender = gender,
    `Blood Type` = blood_type,
    `Medical Condition` = medical_condition,
    `Insurance Provider` = insurance_provider,
    `Admission Type` = admission_type,
    `Admission Date` = date_of_admission,
    `Length of Stay` = length_of_stay,
    `Billing Amount` = billing_amount,
    Medication = medication,
    `Test Result` = test_results
  ) %>%
  slice_head(n = 500)

DT::datatable(
  preview_tbl,
  caption = "De-identified Patient Encounter Preview",
  options = list(pageLength = 10, scrollX = TRUE, autoWidth = TRUE),
  rownames = FALSE,
  filter = "top",
  class = "cell-border stripe"
) %>%
  formatCurrency("Billing Amount", currency = "$", digits = 0) %>%
  formatRound("Length of Stay", digits = 0)
```

## 3. Feature Engineering

Feature engineering creates practical groupings that support clearer interpretation. Here, age is grouped into patient segments, length of stay is grouped into utilization bands, billing amount is split into quartile-based tiers, and admission weekday is created from the admission date.

```{r feature-engineering}
health_clean <- health_clean %>%
  mutate(
    age_group = cut(
      age,
      breaks = c(0, 17, 34, 49, 64, Inf),
      labels = c("0-17", "18-34", "35-49", "50-64", "65+"),
      include.lowest = TRUE
    ),
    length_of_stay_group = cut(
      length_of_stay,
      breaks = c(-Inf, 2, 7, 14, Inf),
      labels = c("0-2 days", "3-7 days", "8-14 days", "15+ days")
    ),
    billing_tier = ntile(billing_amount, 4),
    billing_tier = factor(
      billing_tier,
      levels = 1:4,
      labels = c("Lowest 25%", "25-50%", "50-75%", "Highest 25%")
    ),
    admission_weekday = wday(date_of_admission, label = TRUE, abbr = FALSE)
  )

summary(health_clean)
```

## 4. Executive Summary Metrics

The table below provides a quick profile of the cleaned analytical sample.

```{r executive-summary}
top_condition <- health_clean %>%
  count(medical_condition, sort = TRUE) %>%
  slice_max(n, n = 1, with_ties = FALSE)

top_insurer <- health_clean %>%
  count(insurance_provider, sort = TRUE) %>%
  slice_max(n, n = 1, with_ties = FALSE)

date_range_text <- paste(
  format(min(health_clean$date_of_admission), "%B %d, %Y"),
  "to",
  format(max(health_clean$date_of_admission), "%B %d, %Y")
)

overview_metrics <- tibble(
  Metric = c(
    "Patient encounters analyzed",
    "Admission date range",
    "Median patient age",
    "Median billing amount",
    "Median length of stay",
    "Most common condition",
    "Largest insurance group"
  ),
  Value = c(
    scales::comma(nrow(health_clean)),
    date_range_text,
    paste0(round(median(health_clean$age, na.rm = TRUE), 1), " years"),
    format_currency(median(health_clean$billing_amount, na.rm = TRUE)),
    paste0(round(median(health_clean$length_of_stay, na.rm = TRUE), 1), " days"),
    paste0(as.character(top_condition$medical_condition), " (", format_percent(top_condition$n / nrow(health_clean)), ")"),
    paste0(as.character(top_insurer$insurance_provider), " (", format_percent(top_insurer$n / nrow(health_clean)), ")")
  )
)

knitr::kable(overview_metrics, caption = "Executive Summary Metrics")
```

The cleaned dataset contains `r scales::comma(nrow(health_clean))` patient encounters from `r date_range_text`. The most common condition is `r as.character(top_condition$medical_condition)`, and the largest insurance group is `r as.character(top_insurer$insurance_provider)`.

## 5. Exploratory Data Analysis

Exploratory data analysis summarizes the main distributions and group-level differences before formal testing.

### 5.1 Summary by Medical Condition

```{r condition-summary}
condition_summary <- health_clean %>%
  group_by(medical_condition) %>%
  summarise(
    Patients = n(),
    Patient_Share = Patients / nrow(health_clean),
    Median_Age = median(age, na.rm = TRUE),
    Median_Length_of_Stay = median(length_of_stay, na.rm = TRUE),
    Median_Billing = median(billing_amount, na.rm = TRUE),
    Average_Billing = mean(billing_amount, na.rm = TRUE),
    Abnormal_Test_Rate = mean(test_results == "Abnormal", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(Patients))

DT::datatable(
  condition_summary,
  caption = "Patient and Billing Summary by Medical Condition",
  options = list(pageLength = 10, autoWidth = TRUE),
  rownames = FALSE,
  filter = "top",
  class = "cell-border stripe"
) %>%
  formatPercentage(c("Patient_Share", "Abnormal_Test_Rate"), digits = 1) %>%
  formatRound(c("Median_Age", "Median_Length_of_Stay"), digits = 1) %>%
  formatCurrency(c("Median_Billing", "Average_Billing"), currency = "$", digits = 0)
```

This table helps identify which conditions are most common and whether those conditions differ meaningfully in age profile, stay length, billing, or abnormal test rate.

### 5.2 Summary by Insurance Provider

```{r insurance-summary}
insurance_summary <- health_clean %>%
  group_by(insurance_provider) %>%
  summarise(
    Patients = n(),
    Patient_Share = Patients / nrow(health_clean),
    Median_Billing = median(billing_amount, na.rm = TRUE),
    Average_Billing = mean(billing_amount, na.rm = TRUE),
    Median_Length_of_Stay = median(length_of_stay, na.rm = TRUE),
    Emergency_Share = mean(admission_type == "Emergency", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(Patients))

DT::datatable(
  insurance_summary,
  caption = "Billing and Utilization Summary by Insurance Provider",
  options = list(pageLength = 10, autoWidth = TRUE),
  rownames = FALSE,
  filter = "top",
  class = "cell-border stripe"
) %>%
  formatPercentage(c("Patient_Share", "Emergency_Share"), digits = 1) %>%
  formatRound("Median_Length_of_Stay", digits = 1) %>%
  formatCurrency(c("Median_Billing", "Average_Billing"), currency = "$", digits = 0)
```

Insurance-level summaries can highlight where patient volume, emergency mix, or billing levels differ across payers.

### 5.3 Admission Type and Condition Mix

```{r admission-condition-table}
admission_condition <- health_clean %>%
  count(medical_condition, admission_type) %>%
  group_by(medical_condition) %>%
  mutate(Condition_Share = n / sum(n)) %>%
  ungroup() %>%
  pivot_wider(
    names_from = admission_type,
    values_from = c(n, Condition_Share),
    values_fill = 0
  )

DT::datatable(
  admission_condition,
  caption = "Admissions by Medical Condition and Admission Type",
  options = list(pageLength = 10, scrollX = TRUE, autoWidth = TRUE),
  rownames = FALSE,
  class = "cell-border stripe"
)
```

This cross-tabulation shows whether each condition has a different mix of elective, urgent, and emergency admissions.

## 6. Visual Analysis

Interactive visualizations make it easier to explore billing variation, utilization patterns, and potential relationships among key variables.

### 6.1 Billing Amount Distribution

```{r billing-distribution}
median_billing <- median(health_clean$billing_amount, na.rm = TRUE)

p_billing <- ggplot(health_clean, aes(x = billing_amount)) +
  geom_histogram(bins = 40, fill = "#2c7fb8", color = "white") +
  geom_vline(xintercept = median_billing, color = "#d95f02", linewidth = 1) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Distribution of Billing Amount",
    subtitle = paste("Median billing amount:", format_currency(median_billing)),
    x = "Billing Amount",
    y = "Patient Encounters"
  )

ggplotly(p_billing)
```

The distribution gives a quick view of common billing ranges and highlights whether high-cost outliers are present.

### 6.2 Billing by Medical Condition

```{r billing-by-condition}
p_condition_billing <- health_clean %>%
  mutate(medical_condition = forcats::fct_reorder(medical_condition, billing_amount, .fun = median)) %>%
  ggplot(
    aes(
      x = medical_condition,
      y = billing_amount,
      fill = medical_condition,
      text = paste(
        "Condition:", medical_condition,
        "<br>Billing:", format_currency(billing_amount),
        "<br>Admission:", admission_type,
        "<br>Length of stay:", length_of_stay, "days"
      )
    )
  ) +
  geom_boxplot(outlier.alpha = 0.18) +
  coord_flip() +
  scale_y_continuous(labels = scales::dollar_format()) +
  guides(fill = "none") +
  labs(
    title = "Billing Amount by Medical Condition",
    x = "Medical Condition",
    y = "Billing Amount"
  )

ggplotly(p_condition_billing, tooltip = "text")
```

Box plots compare median billing and spread by condition, making it easier to spot categories with more variable or higher-cost encounters.

### 6.3 Admission Type Mix by Medical Condition

```{r admission-mix-plot}
admission_mix <- health_clean %>%
  count(medical_condition, admission_type) %>%
  group_by(medical_condition) %>%
  mutate(Patient_Share = n / sum(n)) %>%
  ungroup()

p_admission_mix <- admission_mix %>%
  mutate(medical_condition = forcats::fct_reorder(medical_condition, n, .fun = sum)) %>%
  ggplot(
    aes(
      x = medical_condition,
      y = Patient_Share,
      fill = admission_type,
      text = paste(
        "Condition:", medical_condition,
        "<br>Admission type:", admission_type,
        "<br>Patients:", scales::comma(n),
        "<br>Share:", format_percent(Patient_Share)
      )
    )
  ) +
  geom_col(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Admission Type Mix by Medical Condition",
    x = "Medical Condition",
    y = "Share of Admissions",
    fill = "Admission Type"
  )

ggplotly(p_admission_mix, tooltip = "text")
```

This chart shows whether conditions have similar admission profiles or whether some are more concentrated in emergency, urgent, or elective encounters.

### 6.4 Billing by Age and Admission Type

```{r age-billing-plot}
set.seed(123)
plot_sample <- health_clean %>%
  slice_sample(n = min(5000, nrow(health_clean)))

p_age_billing <- ggplot(plot_sample, aes(x = age, y = billing_amount, color = admission_type)) +
  geom_point(
    aes(
      text = paste(
        "Age:", age,
        "<br>Billing:", format_currency(billing_amount),
        "<br>Condition:", medical_condition,
        "<br>Admission:", admission_type
      )
    ),
    alpha = 0.35
  ) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Billing Amount by Age and Admission Type",
    subtitle = "A sampled view keeps the interactive chart responsive",
    x = "Age",
    y = "Billing Amount",
    color = "Admission Type"
  )

ggplotly(p_age_billing, tooltip = "text")
```

The sampled scatter plot provides a readable view of whether age is visibly related to billing and whether that relationship differs by admission type.

### 6.5 Length of Stay and Billing

```{r length-of-stay-billing}
p_los_billing <- ggplot(plot_sample, aes(x = length_of_stay, y = billing_amount, color = admission_type)) +
  geom_point(
    aes(
      text = paste(
        "Length of stay:", length_of_stay, "days",
        "<br>Billing:", format_currency(billing_amount),
        "<br>Condition:", medical_condition,
        "<br>Admission:", admission_type
      )
    ),
    alpha = 0.35
  ) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Billing Amount by Length of Stay",
    x = "Length of Stay (days)",
    y = "Billing Amount",
    color = "Admission Type"
  )

ggplotly(p_los_billing, tooltip = "text")
```

Length of stay is a practical utilization measure. If longer stays are associated with higher billing, that can motivate deeper review of care pathways, discharge planning, and case complexity.

### 6.6 Admissions by Day of Week

```{r weekday-admissions}
weekday_admissions <- health_clean %>%
  count(admission_weekday)

p_weekday <- weekday_admissions %>%
  ggplot(aes(x = admission_weekday, y = n, text = paste("Admissions:", scales::comma(n)))) +
  geom_col(fill = "#4daf4a") +
  labs(
    title = "Hospital Admissions by Day of Week",
    x = "Admission Weekday",
    y = "Patient Encounters"
  )

ggplotly(p_weekday, tooltip = "text")
```

Weekday patterns can help operations teams understand whether patient volume is evenly distributed or concentrated around certain days.

## 7. Statistical Analysis

The statistical tests below add structure to the exploratory findings. Because the dataset is observational, results should be interpreted as associations rather than causal effects.

### 7.1 Chi-Square Test: Medical Condition vs. Admission Type

```{r chi-square-test}
condition_admission_tab <- table(health_clean$medical_condition, health_clean$admission_type)
chi_condition_admission <- chisq.test(condition_admission_tab)

chi_summary <- tibble(
  Test = "Chi-square test of independence",
  Statistic = round(unname(chi_condition_admission$statistic), 2),
  Degrees_of_Freedom = unname(chi_condition_admission$parameter),
  P_Value = format_p_value(chi_condition_admission$p.value)
)

knitr::kable(chi_summary, caption = "Association Between Medical Condition and Admission Type")
```

This test evaluates whether admission type is independent of medical condition. A small p-value suggests that admission type mix differs by condition.

### 7.2 Correlation: Age and Billing Amount

```{r age-billing-correlation}
age_billing_cor <- cor.test(
  health_clean$age,
  health_clean$billing_amount,
  use = "complete.obs"
)

cor_summary <- tibble(
  Metric = c("Correlation coefficient", "P-value"),
  Value = c(
    round(unname(age_billing_cor$estimate), 3),
    format_p_value(age_billing_cor$p.value)
  )
)

knitr::kable(cor_summary, caption = "Correlation Between Age and Billing Amount")
```

The correlation coefficient describes the strength and direction of the relationship between patient age and billing amount.

### 7.3 Multivariable Billing Model

```{r billing-regression}
billing_model <- lm(
  billing_amount ~ age + length_of_stay + admission_type +
    medical_condition + insurance_provider,
  data = health_clean
)

model_fit <- summary(billing_model)
model_f <- model_fit$fstatistic
model_p <- pf(model_f[1], model_f[2], model_f[3], lower.tail = FALSE)

model_summary <- tibble(
  Metric = c("R-squared", "Adjusted R-squared", "Overall model p-value"),
  Value = c(
    round(model_fit$r.squared, 3),
    round(model_fit$adj.r.squared, 3),
    format_p_value(model_p)
  )
)

knitr::kable(model_summary, caption = "Billing Model Fit Summary")
```

```{r billing-regression-coefficients}
model_coefficients <- as.data.frame(coef(summary(billing_model))) %>%
  tibble::rownames_to_column("Term") %>%
  rename(
    Estimate = Estimate,
    Std_Error = `Std. Error`,
    Statistic = `t value`,
    P_Value = `Pr(>|t|)`
  ) %>%
  filter(Term != "(Intercept)") %>%
  mutate(Absolute_Estimate = abs(Estimate)) %>%
  arrange(desc(Absolute_Estimate)) %>%
  slice_head(n = 10) %>%
  select(Term, Estimate, Std_Error, Statistic, P_Value)

DT::datatable(
  model_coefficients,
  caption = "Largest Billing Model Coefficients by Absolute Estimate",
  options = list(pageLength = 10, autoWidth = TRUE),
  rownames = FALSE,
  class = "cell-border stripe"
) %>%
  formatCurrency(c("Estimate", "Std_Error"), currency = "$", digits = 0) %>%
  formatRound("Statistic", digits = 2) %>%
  formatSignif("P_Value", digits = 3)
```

The regression model provides a directional view of billing associations after controlling for age, length of stay, admission type, medical condition, and insurance provider. It should be used as an exploratory diagnostic rather than a final predictive model.

## 8. Conclusion and Recommendations

This report cleaned and structured the healthcare dataset, removed non-positive billing records, created utilization-focused features, and analyzed billing variation across conditions, insurance providers, admission types, patient age, and length of stay. The updated workflow is designed to be reproducible, privacy-aware, and easier to present in a professional portfolio.

Key recommendations:

- Use condition and insurance summaries to identify groups with the highest median billing and largest patient volumes.
- Review admission type mix by condition to understand whether certain diagnoses are more concentrated in emergency, urgent, or elective encounters.
- Treat age, length of stay, and admission type as starting points for deeper cost-driver analysis, not as causal explanations.
- Keep public-facing outputs de-identified by excluding direct patient, provider, room, and facility identifiers from preview tables.

Future work could extend this project with predictive modeling, train/test validation, feature importance, and a dashboard-style summary for hospital operations or insurance analytics.
